OpenMPIR

### 2017-09-08 paper.tex
Rework Figure 5 to be consistent with other code figures
Unnecessary capitalization
More clean up suggestions.

For the cases where performance is close between the implementations, like
\texttt{sparselu}, we expect runtimes to be dominated by actual task work. This 
expectation is confirmed by investigating performance counters for \texttt{sparselu}:
every implementation spends roughly 90\% of it's time doing work in the task
bodies.

@@

In this section we discuss the relevance of the results and this work in general
in the context of the literature. As discussed in Section~\ref{Sec:Introduction}, 
there have been many programming models for implementing parallel programs. While
this paper does not take a stance on which one \emph{should} be used, it does
accept that OpenMP is currently a widely used model, potentially the most
widely used model that has a notion of tasks.

The goal of this work is to show that a single intermediate representation can be
used to implement multiple programming models, with good performance. Schardl et 
al. showed that Tapir works well as an IR for Cilk, and surmised that it could
be an effective IR for a large subset of OpenMP \cite{tapir}. This work has
taken the first step in that direction, showing that it can work well as an IR
for OpenMP tasks.

One interesting aspect of this work is that it compiles a language extension intended
for one runtime (OpenMP runtimes) to use another (the Cilk runtime). While in many
ways this is a weakness of this work, it does open the possibility of doing this 
in a general way. Having a real IR to work with enables, in the same way that LLVM
does for serial code, the possibility of multiple input parallel languages and
multiple output parallel runtimes. While the OpenMP spec's requirement to
have access to runtime routines complicates things, for many programming models
it seems at least conceivable to map them onto multiple different runtimes. This
opens the possibility of mixing programming models at compile time by targeting
the same runtime. For example, if both OpenMP and Cilk targeted Tapir
instructions, one could have a large application where some parts were written
using Cilk, while other used OpenMP, and the result would not suffer the
incompatibility problems often seen today.

The surprising fact that program behaviour wasn't changed significantly is
worth discussing here. We surmise that it is likely due to the variable
semantics borrowed from the Cilk codegen and their relation to the OpenMP
specification of semantics. In particular, \emph{OpenMP doesn't specify any
memory model for shared variables}. This results in the Cilk style of only
writing to a shared variable at the end of a parallel section as a perfectly
valid implementation of OpenMP's specification. In contrast, existing OpenMP
implementations often write to shared variables on every write. This is a
potential for performance discrepancy due to differing implementation-defined
behaviour, and we will return to it when discussing performance on the
Barcelona OpenMP task suite. There is of course the fact that technically, by
copying the value of variables back to the surrounding context even for
\texttt{private} variables, one isn't following the specification. We surmise
that in practice this hasn't had an effect due to the fact that
\texttt{private} variables are generally used for performance, rather than
their semantic properties.

>>>>

For the cases where performance is close between the implementations, like
\texttt{sparselu}, we expect runtimes to be dominated by actual task workloads. This 
expectation is confirmed by investigating performance counters for \texttt{sparselu}
where every implementation spends roughly 90\% of it's time in the task
bodies.

@@

In this section we discuss the relevance of the results and in the context of the
literature. As discussed in Section~\ref{Sec:Introduction}, 
there have been many programming models for implementing parallel programs. While
this paper does not take a stance on which one \emph{should} be used, it does
accept that OpenMP is currently/potentially the most widely used model that has 
a notion of tasks. \pmnote{but it isn't clear to me that tasks are really being 
used much! :-)}.  The goal of this work is to show that a single intermediate representation can be
used to implement multiple programming paradigms with good performance. Schardl et 
al. showed that Tapir works well as an IR target for Cilk and surmised that it could
be an effective IR for a large subset of OpenMP \cite{tapir}. This work has
taken the first step in that direction.

One interesting aspect of this work is that it compiles a language extension intended
for one runtime (OpenMP) to use another (the Cilk runtime). While in many
ways this is a weakness of this work, it does open the possibility of utilizing this 
in a more general way to address many aspects including significant challenges 
such as performance portability. Having a parallel-aware IR to work with enables, 
in the same way that LLVM does for serial code, the possibility of multiple high-level 
parallel languages and multiple parallel runtime targets. While the OpenMP specification's 
requirement to have access to runtime routines complicates things, for many programming models
it seems at least conceivable to map them onto multiple different runtimes. This
opens the possibility of mixing programming models at compile time by targeting
the same runtime. For example, if both OpenMP and Cilk targeted Tapir
instructions, one could have a large application where some parts were written
using Cilk, while other used OpenMP, and the result would not suffer the
incompatibility problems often seen today.

The surprising fact that program behaviour wasn't changed significantly is
worth discussing here. We surmise that it is likely due to the variable
semantics borrowed from the Cilk codegen and their relation to the OpenMP
specification of semantics. In particular, \emph{OpenMP doesn't specify a
memory model for shared variables}. This results in the Cilk style of only
writing to a shared variable at the end of a parallel section as a perfectly
valid implementation of OpenMP's specification. In contrast, existing OpenMP
implementations often write to shared variables on every write. This is a
potential for performance discrepancy due to differing implementation-defined
behaviour and we will return to it when discussing performance on the
Barcelona OpenMP task suite. There is of course the fact that technically, by
copying the value of variables back to the surrounding context even for
\texttt{private} variables, one isn't following the specification. We surmise
that in practice this hasn't had an effect due to the fact that
\texttt{private} variables are generally used for performance, rather than
their semantic properties.
